---
title: "Cleaning pipeline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cleaning pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup, message = FALSE}
library(concert)
```

This vignette covers the offline half of the pipeline: reading a messy file,
finding the header row, tagging columns, and cleaning names and CAS numbers.
Nothing here touches the network.

## Read a file with frontmatter

Lab exports usually start with report metadata. Here is a small CSV with four
lines of frontmatter before the real header, padded with commas the way Excel
writes them.

```{r}
csv_path <- tempfile(fileext = ".csv")
writeLines(c(
  "Chemical Inventory Report,,,,",
  "Generated: 2024-01-15,,,,",
  "Laboratory: Organic Chemistry Lab,,,,",
  ",,,,",
  "Chemical Name,CAS Number,Result,Unit,Location",
  "Acetone,67641,2.5,mg/L,Building A",
  "  Ethanol  ,64-17-5,< 0.5,ug/L,Building A",
  "Toluene,108-88-3,12,mg/L,Building B",
  "Lead + Zinc,,0.3,ppb,Building B"
), csv_path)

raw <- safely_read_file(csv_path, "csv")
raw
```

`detect_data_start()` runs three detectors (fill-ratio heuristic, chemistry
keyword matching, type consistency) and keeps the most confident answer.

```{r}
detection <- detect_data_start(raw)
detection[c("header_row", "data_start_row", "method", "confidence")]
```

If detection is wrong, pass the row yourself:

```r
detection <- detect_data_start(raw, mode = "manual", manual_row = 5)
```

`extract_clean_data()` promotes the header row and drops everything above it.
The app then runs `handle_merged_cells()` and `janitor::clean_names()`, so do
the same here.

```{r}
clean <- extract_clean_data(raw, detection)
clean <- janitor::clean_names(handle_merged_cells(clean))
clean
```

## Tag columns

Tags tell the pipeline what each column means. `suggest_column_tags()`
guesses from header names only, and stays quiet when unsure.

```{r}
suggest_column_tags(names(clean))
```

Suggestions are a starting point. Build the final `tag_map` yourself; every
`Result` column needs a `Unit` partner.

```{r}
tag_map <- list(
  chemical_name = "Name",
  cas_number = "CASRN",
  result = "Result",
  unit = "Unit"
)
is.null(validate_tag_pairing(tag_map))  # NULL means no pairing problem
```

## Run the cleaning pipeline

`run_cleaning_pipeline()` chains the cleaning steps and records every change.

```{r}
result <- run_cleaning_pipeline(clean, tag_map)
result$cleaned_data
```

The audit trail has one row per changed cell: which step touched it and the
before and after values. Here the CAS for acetone was normalized to the
hyphenated form, and the multi-analyte cell was flagged.

```{r}
result$audit_trail[, c("row_id", "field", "step", "original_value", "new_value")]
```

The `Lead + Zinc` row was flagged, not split. Multi-analyte cells are left for
review because the right action (split, keep, rename) depends on the dataset.
See `resolve_multi_analyte_row()` and the headless vignette for how to resolve
them in bulk.

### Switch steps off

Some datasets are mangled by a step that helps elsewhere. `mask` turns
individual steps off. Names are `unicode`, `whitespace`, `cas`, `names`,
`isotopes`, `multi`, `chiral`, `truncated`, `bare_formula`, and
`reference_flags`.

```{r}
no_multi <- run_cleaning_pipeline(clean, tag_map, mask = list(multi = FALSE))
unique(no_multi$audit_trail$step)
```

### Fix values before cleaning

`apply_value_corrections()` runs column-scoped pattern replacements before
the pipeline. Use it for known prefixes, suffixes, and typos.

```{r}
corrections <- data.frame(
  column = "chemical_name",
  pattern = "^Total ",
  replacement = ""
)
fixed <- apply_value_corrections(
  data.frame(chemical_name = c("Total Lead", "Zinc")),
  corrections
)
fixed$cleaned_data
```

## CAS helpers

The CAS utilities are exported for use outside the pipeline.

```{r}
is_cas(c("67-64-1", "67-64-2", "not a cas"))
as_cas(c("67641", "64-17-5"))
extract_cas("ethanol (64-17-5) in water")
```

## Reference lists

The name-cleaning steps read bundled reference lists: stop words, strip terms,
block patterns, functional categories, and an isotope lookup. Load them to
inspect or extend.

```{r, message = FALSE}
refs <- load_all_reference_lists()
names(refs)
head(refs$strip_terms, 3)
```

User edits persist through `save_user_reference_lists()` and are merged over
the package defaults on the next load.
