---
title: "Measurement harmonization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Measurement harmonization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup, message = FALSE}
library(concert)
```

Harmonization turns free-text results, unit strings, and media labels into
numeric values in ToxVal target units. In `curate_headless()` it runs when
`harmonize = TRUE`; in the app it is the Harmonize tab. The building blocks
are exported and work on plain vectors.

## Parse numeric results

Result columns mix numbers with qualifiers, detection limits, and
scientific notation in several spellings.

```{r}
parsed <- parse_numeric_results(c("< 5.0", "2.5x10^3", "BDL", "4.56+02", "≥100", "5-10", "1e-3"))
parsed
```

Each row keeps the original string, the parsed value, and a qualifier or
flag explaining anything that was not a plain number.

## Harmonize units

`harmonize_units()` converts values into the target unit for each unit
family using the bundled conversion table.

```{r, message = FALSE}
unit_map <- load_unit_map()
head(unit_map[, c("from_unit", "to_unit", "multiplier")])
```

```{r}
harmonize_units(
  values = c(2.5, 500, 12, 0.3),
  units = c("mg/L", "ug/L", "µg/l", "ppb"),
  unit_map = unit_map
)
```

Notes on the lookup:

- Micro symbols and whitespace are normalized before lookup.
- Exact, case-sensitive matches win. Case-insensitive fallback only applies when
  every candidate row agrees on the target and multiplier.
- Unknown units pass through unchanged with `unit_flag = "unmatched"`.

### ppb and ppm need a media context

`ppb` in water means micrograms per litre. In soil it means micrograms per
kilogram. In air it is a volume ratio that cannot be converted without
molecular weight, temperature, and pressure. Pass `media` to route these.

```{r}
harmonize_units(
  values = c(1, 1, 1),
  units = c("ppb", "ppb", "ppb"),
  unit_map = unit_map,
  media = c("aqueous", "solid", "air")
)[, c("orig_unit", "harmonized_value", "harmonized_unit", "unit_flag")]
```

### Molarity needs a molecular weight

Units like `mM` or `mol/L` convert to mass concentration only with a
molecular weight. Supply it directly, or pass `dtxsid` and let the function
look it up from CompTox (network required).

```{r}
harmonize_units(
  values = 1,
  units = "mM",
  unit_map = unit_map,
  molecular_weight = 58.08
)[, c("orig_unit", "harmonized_value", "harmonized_unit", "unit_flag")]
```

## Harmonize media

Media labels are mapped to canonical terms with ENVO identifiers and a
routing category (`aqueous`, `air`, `solid`). That category is what feeds
the ppb/ppm routing above.

```{r}
harmonize_media(c("surface water", "Soil", "ambient air", "mystery matrix"))[
  , c("raw_media", "canonical_media", "envo_id", "media_category", "media_flag")
]
```

Unmatched labels get `media_flag = "media_unmatched"` and no route. Add
them to the media map in the app's Harmonize tab, or pass a custom
`media_map`.

## Dates and uncertainty

```{r}
parse_dates(c("2024-01-15", "01/15/2024", "15 Jan 2024", "not a date"))
```

Uncertainty columns are normalized to two-sigma with
`normalize_uncertainty_to_two_sigma()`, using the tagged
`UncertaintyCoverage` column when present.

## Putting it together

`curate_headless(harmonize = TRUE)` runs these steps on the tagged columns,
then `map_to_toxval_schema()` assembles the 56-column ToxVal table. The
returned list carries the harmonization audit alongside the data:

```r
out <- curate_headless(
  input_path = "input.xlsx",
  output_path = "output.xlsx",
  tag_map = list(chemical_name = "Name", casrn = "CASRN", result = "Result", unit = "Unit", matrix = "Media"),
  harmonize = TRUE,
  format = "both"
)
out$data              # ToxVal rows
out$harmonize_audit   # one row per converted or flagged value
```
