---
title: "Headless curation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Headless curation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Headless curation runs the whole pipeline from a script. It queries the
CompTox dashboard, so the code in this vignette is shown but not executed.

## One call

```{r}
library(concert)

out <- curate_headless(
  input_path = "data/inventory.xlsx",
  output_path = "output/inventory_curated.xlsx",
  tag_map = list(
    chemical_name = "Name",
    cas_number = "CASRN",
    result = "Result",
    unit = "Unit"
  )
)
```

`tag_map` keys are column names after `janitor::clean_names()`, so
`"CAS Number"` becomes `cas_number`. Run `suggest_column_tags()` on the
cleaned names if you are unsure (see `vignette("cleaning-pipeline")`).

The output workbook has one sheet per pipeline stage: cleaned data,
resolution state, cleaning audit, consensus summary, and a `Pipeline Config`
sheet that lets the app or `parse_concert_export()` reload the run.

The return value mirrors the workbook:

```{r}
out$data          # one row per input row, with dtxsid, preferred name, and status
out$audit_trail   # every cleaning change
```

Set `write_files = FALSE` to get the same list without touching disk.

## Useful options

| Argument | Use it when |
|---|---|
| `header_row` | Frontmatter detection picked the wrong row. |
| `value_corrections` | A prefix, suffix, or typo affects many rows. Applied before cleaning. |
| `cleaning_steps` | A cleaning step mangles this dataset. `list(chiral = FALSE)` and so on. |
| `postprocess_candidates` | You want the same enrichment, scoring, and auto-resolve pass as the app. |
| `accept_suggestions` | Accept every candidate scored `suggested` after postprocessing. |
| `review_picks` | You know the right DTXSID for specific names. Validated against CompTox. |
| `row_flags` | Mark rows `FOLLOW-UP`, `BAD`, or `VERIFIED` with a reason. |
| `multi_analyte_resolutions` | Split, keep, or rename cells like `"Lead + Zinc"`. |
| `harmonize`, `media`, `format` | Parse results, convert units, and write ToxVal parquet or CSV. |
| `starts_with` | Allow a CompTox starts-with fallback for names nothing else matched. |

Review inputs are content-keyed. A `review_picks` row matches on the cleaned
`name` (and `casrn` when given), never on row position, so decisions survive
re-runs after upstream corrections change the row order.

```{r}
out <- curate_headless(
  input_path = "data/inventory.xlsx",
  output_path = "output/inventory_curated.xlsx",
  tag_map = list(chemical_name = "Name", cas_number = "CASRN"),
  postprocess_candidates = TRUE,
  accept_suggestions = TRUE,
  review_picks = data.frame(
    name = "TKN",
    dtxsid = "DTXSID0020000"
  ),
  row_flags = data.frame(
    name = "Total PFAS",
    flag = "FOLLOW-UP",
    reason = "Group total, no single DTXSID"
  ),
  value_corrections = data.frame(
    column = "chemical_name",
    pattern = "^Dissolved ",
    replacement = ""
  )
)
```

## Stages

`curate_headless()` is a composition of six stage functions. Call them
yourself when you want to stop partway, inspect state, or swap a step.

```{r}
state <- stage_ingest(input_path = "data/inventory.xlsx", tag_map = tag_map)
state <- stage_clean(state)
state$cleaning_result$audit_trail

state <- stage_curate(state)
state$consensus_summary

state <- stage_review(state, accept_suggestions = TRUE)
state <- stage_harmonize(state, harmonize = TRUE)
stage_export(state, output_path = "output/inventory_curated.xlsx")
```

Each stage adds fields to the same state list. The header comment in
`R/curate_stages.R` lists them.

## Replay scripts

`generate_concert_script()` writes a standalone script that reproduces a run
with one `curate_headless()` call. It embeds every decision plus compact
snapshots of the reference lists, unit map, and media map, so the replay is
pinned even after package defaults change.

```{r}
script <- generate_concert_script(
  input_path = "data/inventory.xlsx",
  output_path = "output/inventory_curated.xlsx",
  tag_map = tag_map,
  header_row = state$detection$header_row,
  reference_lists = state$reference_lists,
  accept_suggestions = TRUE,
  review_picks = data.frame(name = "TKN", dtxsid = "DTXSID0020000")
)
writeLines(script, "output/replay.R")
```

The app's export does the same from the Review Results tab, and the agent
loop below writes one on every run. Commit `replay.R` with the data. It is
the reproducibility record.

## Agent loop

For unattended curation, the package ships a decide-run-inspect loop. An
agent (or you) edits one file, `decisions.R`, and re-runs until no row is
pending.

```{r}
curate_decisions_template("data/inventory.xlsx", out_dir = "output/loop")
```

This writes `output/loop/decisions.R` with the detected header row, suggested
tags, a pinned reference-list snapshot, and a commented schema for every other
decision object. Then:

```{r}
curate_iterate("output/loop/decisions.R")
```

Each run writes:

- `status.md`: done or not, errors, unmatched decision keys, and a list of
  low-similarity matches that deserve a second look.
- `pending.csv`: one row per unresolved data row with a `pending_type`
  (`multi_analyte`, `verified_unresolved`, `candidate_validation`,
  `identity_scope`, `disagree`, `suggested`, `no_match`), the suggested
  DTXSID or split, and the candidate list.
- `replay.R`: the standalone reproduction script.
- `cache/`: CompTox search results, so re-runs only re-query changed names.

Resolve pending rows by adding to the decision objects in `decisions.R`
(`review_picks`, `row_flags`, `multi_analyte_resolutions`,
`value_corrections`, `cleaning_steps`, reference-list overrides), then
re-run. Root-cause fixes that clear many rows beat per-row picks.

A command-line runner and an agent skill document the same loop:

```{r}
system.file("scripts/curate_loop.R", package = "concert")
system.file("skills/concert-curate/SKILL.md", package = "concert")
```

```bash
Rscript <runner> --template input.csv output/loop
Rscript <runner> output/loop/decisions.R   # exit 0 = done, exit 2 = rows pending
```


## Review evidence and accepted identities

`done` and `queue_complete` describe queue disposition. A retained FOLLOW-UP can
complete that queue while `reconciliation_complete`, `identity_review_complete`
or `review_complete` remain false. Each iteration writes `review_reconciliation.csv`,
`candidate_review.csv`, `identifier_diagnostics.csv` and `identity_review.csv`.
Changed evidence can reopen candidate validation; repeating a VERIFIED flag cannot
resolve a blank/tied identity. Reviewed WQX canonical names retain their separate
name-only contract and never manufacture a DTXSID.

Historical flags have no decision snapshot. They report `baseline_missing`,
regardless of the reason's wording. To record evidence actually inspected at an
explicit decision, use the shared portable object:

```{r eval=FALSE}
run <- curate_iterate("output/loop/decisions.R")
review_decision_evidence <- capture_review_state(
  run$state, name = "Reviewed analyte", disposition = "deferred",
  flag = "FOLLOW-UP", reason = "Specific source-scope question remains"
)
```

Save that object in `decisions.R`; generated replay and workbook Session State
preserve its immutable revisions. `acknowledge_review_evidence()` binds an explicit
acknowledgment to the decision revision, exact source/content scope and current
evidence fingerprint. It resolves that reconciliation event, without clearing
FOLLOW-UP or accepting an identity. A later evidence or scope change reopens work.
Use `validate_review_candidates()` with an explicit configured validator to
recheck saved candidates independently of name discovery. It returns structured
outcomes and a versioned cache; outages are retryable, and no DSSTox download is
triggered. Candidate ordering, duplicates and audit timestamps are not new evidence.

The `DTXSID` tag supplies attributed source evidence. Registry membership is
validated independently of source-identity correspondence; successful source,
resolver and PubChem agreement does not promote a source ID. Untagged ID columns
warn, while deliberate metadata can be retained without queries:

```{r eval=FALSE}
curate_headless(
  input_path = "input.csv", output_path = "curated.xlsx",
  tag_map = list(chemical_name = "Name", source_dtxsid = "DTXSID")
)
# Alternatively omit source_dtxsid from tag_map and retain it only as metadata:
# ignored_identifier_cols = "source_dtxsid"
```

Ignored columns must have no chemical evidence role. Existing unignored Other
roles retain their generic-search behavior. Raw input columns named `dtxsid` or
`dtxsid_*` cannot bypass the generated-evidence registry.

Lookup consensus remains in Curated Data. `identity_review_state()` derives
blockers and `accepted_dtxsid`; `accepted_identity_view()` retains eligible rows
with all source lineage. The Accepted Identities sheet is additive. Registered
mixtures can qualify through explicit correspondence review; aggregates may be
kept unresolved without selecting a representative component. Split rows retain
original name/CAS and part lineage, with repeated component CAS blocked until
explicit mapping or correspondence review.

Structured acceptance is available through `identity_decisions` and
`apply_identity_decisions()`; the GUI displays blockers and source evidence.
Each record requires a unique source/content selector plus original lineage,
resolved scope/conflict, reason, evidence reference, selected ID, explicit
`correspondence = TRUE`, and `identity_evidence_fingerprint()` from the inspected
row. Registry membership is checked separately. Changed or ambiguous evidence
fails closed. Name/CAS selectors may span many rows for flags and ordinary picks;
they cannot express source-specific scope acceptance.

For safe ToxVal output, opt in to `toxval_identity_mode = "accepted"` in headless
or iteration decisions when `harmonize = TRUE`. Every measurement row remains,
with NA identifiers for blocked identities and no raw-ID fallback. Existing
lookup defaults are retained as an explicit compatibility choice; this audit mode
is not an accepted-identity filter. The standalone mapper supports the equivalent
`identity_mode = "accepted"`. Workbook import, GUI refresh and generated replay
preserve the explicit mode; accepted-only refresh blanks IDs if current alignment
cannot be verified.
